---
title: "Decision boundaries: KNN vs Naïve Bayes"
author: "Stu Field"
format:
  gfm:
    preview-mode: raw
    columns: 70
editor_options:
  chunk_output_type: console
---



```{r}
#| label: setup
#| include: FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "figures/knn-bayes-"
)

options(keep.source = TRUE)
library(helpr)
library(libml)
library(withr)
library(ggplot2)
library(patchwork)

sim_data1 <- function(n = 200) {
  withr::local_seed(1001)
  df   <- data.frame(x1 = rnorm(n, 45, 2), x2 = rnorm(n, 75, 2))
  df$y <- factor(ifelse(df$x1 < 44 | df$x2 <= 74, "control", "disease"))
  df
}

sim_data2 <- function(n = 100) {
  withr::local_seed(999)
  df1 <- data.frame(x1 = rnorm(n, 46, 1.5), x2 = rnorm(n, 75, 1.5)) # control
  df2 <- data.frame(x1 = rnorm(n, 44, 1), x2 = rnorm(n, 78, 1))     # disease
  df  <- rbind(df1, df2)
  df$y <- factor(rep(c("control", "disease"), each = n))
  df
}

plot_knn_boundary <- function(data, k = 15L, res = 50L) {
  stopifnot(ncol(data) == 3L)
  train <- dplyr::rename_if(data, is.factor, function(.x) "class") |>
    dplyr::rename_with(function(.x) c("F1", "F2"), !"class")
  df <- libml:::expand_grid(
    list(
      F1 = seq(min(train$F1), max(train$F1), length = res),
      F2 = seq(min(train$F2), max(train$F2), length = res))
    )
  m <- libml::fit_kknn(class ~ ., train = train, test = train,
                       k_neighbors = k)
  pred <- libml::calc_predictions(m, df)
  df$Pr <- pred$prob_disease
  col_palette <- libml:::col_palette
  pos_space <- round(sum(matrix(df$Pr, nrow = res) >= 0.5) / res^2, 3L)
  p <- ggplot(df, aes(x = F1, y = F2))
  p + geom_raster(aes(fill = Pr), alpha = 0.5) +
    geom_contour(aes(x = F1,  y = F2, z = Pr), binwidth = 0.501,
                 color = "navy", linewidth = 0.5, linetype = "dashed") +
    scale_fill_gradient(low    = col_palette$lightgreen,
                        high   = col_palette$purple,
                        limits = c(0, 1),
                        name   = "P(pos_class)",
                        breaks = seq(0, 1, 0.25), 
                        guide  = guide_colorbar(order = 2)) +
    geom_point(data = train, aes(x = F1, y = F2, color = class),
               size = 2.5, alpha = 0.5) +
    scale_color_manual(
      values = c(col_palette$lightgreen, col_palette$purple),
       guide = guide_legend(order = 1)
    ) +
    geom_point(data = train, aes(x = F1, y = F2), size = 2.5,
               shape = 21, color = "black") +
    labs(x = "Feature 1", y = "Feature 2",
         caption = "dashed line: P = 0.5",
         title = sprintf("KNN (k=%i | space = %0.2f)",
                         k, pos_space),
         color = NULL) +
    libml:::libml_theme(legend_pos = "right") +
    NULL
}
```


# Overview

It can be useful to visualize the decision boundaries of various
models. Here we compare 2 commonly used models:

* *k*-nearest neighbors (KNN)
* Naïve Bayes


-----------------




## KNN vs Naïve Bayes

Below are decision boundaries for 2 simulated data sets using
*k*-nearest neighbors and Naïve Bayes models. In the first data
set (upper 2 panels) the true class boundary is simulated such that:

* the disease (purple) Feature_1 \> 44
* Feature_2 \> 74

these data are simulated with an *unrealistic* threshold to form
the classes.



```{r}
#| label: knn-vs-bayes
#| caption: >
#|   Foo.
#| fig.width: 11
#| fig.height: 5
#| echo: FALSE
p1 <- plot_knn_boundary(sim_data1(), k = 15L)
p2 <- libml::plot_bayes_boundary(sim_data1(), pos_class = "disease") +
  labs(title = "Naive Bayes") + theme(legend.position = "none")
p1 + p2
```

The panels below are simulated from bivariate normal distributions,
somewhat more realistic, and show the difference in the boundary
between the two methods.

```{r}
#| label: knn-vs-bayes2
#| caption: >
#|   Bar.
#| fig.width: 11
#| fig.height: 5
#| echo: FALSE
p1 <- plot_knn_boundary(sim_data2(), k = 15L)
p2 <- libml::plot_bayes_boundary(sim_data2(), pos_class = "disease") +
  labs(title = "Naive Bayes") + theme(legend.position = "none")
p1 + p2
```

## Choosing *k* in KNN

Choosing appropriate *k* for the neighborhood can be difficult,
and often arbitrary. It is often useful to plot the decision
boundary at successive values of *k* and visually inspect.

```{r}
#| label: knn-k
#| fig.width: 11
#| fig.height: 10
pk <- lapply(2:10L, plot_knn_boundary, data = sim_data1())
patchwork::wrap_plots(
  pk, ncol = 3L, guides = "collect", axis_titles = "collect"
)
```


---


### Code Reference

```{r}
#| label: strip-env
#| include: false
print.fn <- function(f) {
  withr::local_options(list(width = 120L))
  ch <- nchar(deparse(args(f))[1L])
  cat(strrep("-", ch), "\n")
  environment(f) <- globalenv()    # drops <environment: ...>
  attr(f, "class") <- NULL
  NextMethod()
}
sim_data1 <- add_class(sim_data1, "fn")
sim_data2 <- add_class(sim_data2, "fn")
plot_knn_boundary <- add_class(plot_knn_boundary, "fn")
```
```{r}
#| label: code-ref
#| comment: ""
# simulated dataset #1
sim_data1

# simulated dataset #2
sim_data2

# plot decision boundary for KNN
plot_knn_boundary
```

---

# Session Info {.unnumbered}

```{r}
#| label: get-session
#| include: FALSE
get_session_info <- function(pkgs = NULL) {
  ret <- list(
    packages = dplyr::select(sessioninfo::package_info(), -library),
    platform = sessioninfo::platform_info()
  )
  ret$platform$date <- NULL
  if (!is.null(pkgs)) {
    ret$packages <- ret$packages[pkgs, ]
  }
  ret
}
```
```{r}
#| label: session
#| code-fold: TRUE
get_session_info()
```
