---
title: "Clinical Outcomes and Inverse Probability Weighting"
subtitle: "Adjusting for confounded variables in model fit"
author: "Stu Field"
format:
  gfm:
    preview-mode: raw
    columns: 70
editor_options: 
  chunk_output_type: console
---

::: {.content-visible when-format="gfm"}
*Adjusting for confounded variables in model fit*
:::


```{r}
#| label: setup
#| include: FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = paste0("figures/ipw-")
)
library(dplyr)
library(ggplot2)
library(patchwork)
options(pillar.min_chars = Inf, width = 120L)
calc_ess <- function(w) {
  (sum(w))^2 / sum(w^2)
}
.fmt_pct <- function(x, digits = 1L) {
  sprintf("%.*f%%", digits, 100 * x)
}
trim_weights <- function(x, p = c(0.01, 0.99)) {
  q  <- quantile(x, p = p)
  lo <- unname(q[1L])
  hi <- unname(q[2L])
  x[x > hi] <- hi
  x[x < lo] <- lo
  x
}
colors <- c("#42587A", "#5B0D67", "#00A499")
plot_cdf <- function(data, var, group = NULL, xlim = NULL, ...) {
  var   <- enquo(var)
  group <- enquo(group)
  vars  <- unname(c(var, group, enquos(...)))
  if (rlang::quo_is_null(group)) {
    p <- select(data, !!!vars) |>
      ggplot(aes(x = !!var)) +
      stat_ecdf(colour = colors[1L], linewidth = 0.75,
                linetype = "solid")
  } else {
    p <- select(data, !!!vars) |>
      ggplot(aes(x = !!var, color = !!group, ...)) +
      stat_ecdf(linewidth = 0.75, linetype = "solid") +
      scale_color_manual(values = colors)
  }
  p + coord_cartesian(xlim = xlim) +
    scale_x_continuous(n.breaks = 8L) +
    scale_y_continuous(n.breaks = 6L) +
    labs(x = var, y = bquote(italic(P) ~ (X < x)))
}
plot_pdf <- function(data, var, group = NULL, xlim = NULL,
                     fill = FALSE, linetype = "solid", ...) {

  var   <- enquo(var)
  group <- enquo(group)
  vars  <- unname(c(var, group, enquos(...)))
  if (rlang::quo_is_null(group)) {
    col <- colors[1L]
    p <- select(data, !!!vars) |>
      ggplot(aes(x = !!var, ...)) +
      geom_density(colour = col, fill = ifelse(fill, col, NA),
                   alpha = ifelse(fill, 0.2,  1),
                   linewidth = 0.75, linetype = linetype)
  } else {
    p <- select(data, !!!vars) |>
      ggplot(aes(x = !!var, color = !!group, ...)) +
      geom_density(if (fill) aes(fill = !!group),
                   alpha     = ifelse(fill, 0.2,  1),
                   linewidth = ifelse(fill, 0.25, 0.75),
                   linetype  = linetype) +
      scale_color_manual(values = colors)
    if (fill) {
      p <- p + scale_fill_manual(values = colors)
    }
  }
  p + coord_cartesian(xlim = xlim) +
    scale_x_continuous(n.breaks = 8L) +
    scale_y_continuous(n.breaks = 6L) +
    labs(y = "Probability Density", x = var)
}
```


# Overview

* Typical modeling of IVF outcomes center around probability of
  pregnancy or live birth from an individual ovarian stimulation cycle
* Cumulative life birth rate (CLBR) is defined as the probability
  of *one or more* positive live birth outcomes, from a completed
  stimulation cycle, *regardless* of oocytes retrieved, fertilized
  oocytes, transferred embryos, or clinical pregnancies.
* **analysis goals**: does CLBR differ by treatment (i.e. protocol) type?
  Does the *type* of FSH administered confer a clinical benefit as
  seen in increased CLBR outcomes?


## The Model

We model the *cumulative* live birth outcome per cycle,
$\hat{y}$ = success or failure, as:

$$ y_i \sim \mathrm{Bernoulli}(p_i) $$

where $y_i \in \{0, 1\}$ = outcome for the $i^{th}$ IVF cycle,
$p_i$ = probability of success for the $i^{th}$ IVF cycle,


$$
\log \left(\frac{p_i}{1 - p_i}\right) = \beta_0 +
  \sum_j \beta_j^{\text{prot}}\,\mathbf{1}[\text{protocol}_i = j] +
$$
$$
  \beta_{\text{age}}(\text{age}_i)\ +
$$
$$
  \beta_{\text{wt}}(\text{bmi}_i)\ +
$$
$$
  \sum_k \gamma_k\,\mathbf{1}[\text{region}_i = k]\ +
$$
$$
  \sum_k \gamma_k\,\mathbf{1}[\text{race}_i = k]
  \epsilon_i
$$


## Model Fit in R

```r
model <- stats::glm(
  clbr ~ protocol + age + bmi + census_region + race_census,
  family  = binomial("logit"),
  data    = model_data
)
```

# Confounding Variables

* Real World Data (RWD) is full of confounding/uncontrolled variables
  - live clinical data from the wild
  - it is *not* a fully-balanced clinical trial!
* uncontrolled confounding variables are likely
  - covariates are supposed to b independent, but ...
  - what if *protocol* is confounded by *age*?

::: {style="width: 50%;"}
```{r}
#| label: age-protocol
#| echo: FALSE
data.frame(check.names = FALSE, row.names = "mean age",
  "FSH-a" = 20.5, "FSH-b" = 30.3, "Mix FSH-a + FSH-b" = 40.4) |>
  knitr::kable()
```
:::


$$
clbr \sim protocol\ +\ age\ +\ ...\ +\ \epsilon
$$


# Welcome IPW

* Inverse Probability Weighting
* build a classifier (model) predicting *protocol* given *age*
  - if a strong bias exists, age should predict protocol
  - stronger predictions reflect stronger bias
* new model:

$$
\text{protocol} \sim age\ +\ ...\ +\ \epsilon
$$

* invert the probability
  - $\uparrow$ $P(protocol = protocol_k | X)$ down weighted
  - $\downarrow$ $P(protocol = protocol_k | X)$ up weighted
* aggregate effect $\rightarrow$ create pseudo-population that
  de-couples correlation *protocol* $\sim$ *age*


-----------


# Example in R

```{r}
#| label: calc-weights
#| fig.cap: "Weight distribution by protocol."
#| fig.height: 5
#| fig.width: 10
set.seed(101)
n_vec <- c(500, 1000, 1500)
sd    <- 2
df <- tibble::tibble(
  protocol = factor(rep(c("A", "B", "C"), times = n_vec)),
  age = c(
    rnorm(n_vec[1L], mean = 10, sd = sd),
    rnorm(n_vec[2L], mean = 12, sd = sd),
    rnorm(n_vec[3L], mean = 15, sd = sd)
  )
)

df |>
  group_by(protocol) |>
  summarize(n = n(), mu = mean(age), sigma = sd(age))

# fit 3-class model
model <- nnet::multinom(protocol ~ age, data = df)
probs <- predict(model, type = "probs")

# P(protocol = protocol_k | X)
row_idx <- 1:nrow(df)
col_idx <- match(df$protocol, colnames(probs))
ps      <- probs[cbind(row_idx, col_idx)]

# invert + trim 1%
df$weights <- trim_weights(1 / ps, c(0.01, 0.99))

tapply(df$weights, df$protocol, quantile, p = 0.5)

df |>
  plot_cdf(weights, group = protocol) +
  geom_hline(yintercept = 0.5, alpha = 0.75, linetype = "dashed") +
  labs(title = "CDF distribution of weights by protocol")
```

```{r}
#| label: plot-var
#| fig.cap: "Continuous variable (i.e. age) pre- and post-weighting."
#| fig.height: 5
#| fig.width: 10
#| echo: FALSE
p1 <- df |>
  plot_pdf(age, protocol, fill = TRUE) +
  labs(title = "Unadjusted") +
  guides(fill = "none", color = "none")
p2 <- df |>
  plot_pdf(age, protocol, fill = TRUE, weight = weights) +
  labs(title = "Adjusted")
p1 + p2
```

```{r}
#| label: ess
df |>
  group_by(protocol) |>
  summarize(
    n        = n(),                           # orig sample size
    n_adj    = calc_ess(weights),             # effective sample size
    mean     = mean(age),                        # mu orig
    mean_adj = weighted.mean(age, weights)) |>   # mu adjust
  mutate(prop_n   = .fmt_pct(prop.table(n)),     # prop orig
         prop_adj = .fmt_pct(prop.table(n_adj))) # prop adjust
```


## Final Fit in R

```r
model <- stats::glm(
  clbr ~ protocol,
  weights = weights,
  family  = quasibinomial("logit"),
  data    = model_data
)
```

-----------


# References

* ASA Regulatory [Workshop](https://ww2.amstat.org/meetings/biop/2020/):
[Short Course](https://ww2.amstat.org/meetings/biop/2020/onlineprogram/handouts/SC4-Handouts.pdf)

* **WeightIt package in R**:
[https://ngreifer.github.io/WeightIt/index.html](https://ngreifer.github.io/WeightIt/index.html)

::: {.callout-note}
# Citation
  Greifer N (2026). WeightIt: Weighting for Covariate Balance in
  Observational Studies. doi:10.32614/CRAN.package.WeightIt
  <https://doi.org/10.32614/CRAN.package.WeightIt>.
  R package version 1.7.0, <https://CRAN.R-project.org/package=WeightIt>.
:::



# Session Info {.unnumbered}

```{r}
#| label: get-session
#| include: FALSE
get_session_info <- function(pkgs = NULL) {
  ret <- list(
    packages = dplyr::select(sessioninfo::package_info(), -library),
    platform = sessioninfo::platform_info()
  )
  ret$platform$date <- NULL
  if (!is.null(pkgs)) {
    ret$packages <- ret$packages[pkgs, ]
  }
  ret
}
```
```{r}
#| label: session
#| code-fold: TRUE
get_session_info()
```
